Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-03-20 paper-1/elife.tex
Fixed figsupp labels for uniqueness

    {\includegraphics[width=\linewidth]{fig-threshold-initial-withSTD}}\label{figsupp:withSTD}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{fig-thresholds-follow-std}
    \caption{%
        Lower thresholds in \dev{} are a consequence of lower \dev{} non-target response magnitude.
        \textbf{A} $\T{A}{i}$ in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages in the \dev{} sequence ($\T[dev]{A}{i}$) and \msc{} sequence ($\T[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \T{A}{i}$). A and B stimulus locations are highlighted in the left plot; see \FIG{setup} for the locations of the remaining stimuli used in the \msc{} sequence.
        \textbf{B} Average response (in spikes per trial) to non-target trials ($\R{N}{i}$, with $N$ meaning B in \dev{}, and B, C, D, and E in \msc{}). Columns are arranged as in panel A.
        \textbf{C} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$, and associated linear regression.
        \textbf{D} $\Delta \R{N}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \R{N}{i}$ vs $\Delta \T{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
    }
    \label{fig:thresholds-follow-std}
    % 
    \figsupp[The equivalent figure in the full model with both TA and STD.]{%
        The equivalent of \FIG{thresholds-follow-std} in the full model with both TA and STD.
        \textbf{A} $\T{A}{i}$ in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages in the \dev{} sequence ($\T[dev]{A}{i}$) and \msc{} sequence ($\T[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \T{A}{i}$). A and B stimulus locations are highlighted in the left plot; see \FIG{setup} for the locations of the remaining stimuli used in the \msc{} sequence.
        \textbf{B} Average response (in spikes per trial) to non-target trials ($\R{N}{i}$, with $N$ meaning B in \dev{}, and B, C, D, and E in \msc{}). Columns are arranged as in panel A.
        \textbf{C} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$, and associated linear regression.
        \textbf{D} $\Delta \R{N}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \R{N}{i}$ vs $\Delta \T{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
    }
    {\includegraphics[width=\linewidth]{fig-thresholds-follow-std-withSTD}}\label{figsupp:withSTD}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{fig-early-adaptation}
    \caption{%
        Dev non-target responses are reduced due to adaptation in early responders to B.
        \textbf{A} The sample network, colored by latency rank, i.e. the rank order of the time to first spike in response to stimulation at site B in any sequence. The response develops asymmetrically due to structural properties, not due to interactions with A. Note that non-responding neurons are ranked last and appear dark blue.
        \textbf{B} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{B}{i}$, and associated linear regression. Neurons are colored by latency rank as in panel A. Additionally, excitatory neurons among the earliest 100 responders are shown in the left plot, while later responders are shown in the right plot. Inset histograms show the distributions of the plotted values, with the confidence intervals of their medians, calculated by bootstrap with 10000 iterations, in darker color.
        \textbf{C} Median non-target response contrast $\Delta \R{N}{}$ among the first 100 neurons to respond to stimulation in B, labeled ``early'', and among the later neurons, labeled ``late'', across networks and stimuli.
        \textbf{D} Median contrast in $\T{B}{}$ among the early and late neurons, respectively, across networks and stimuli.
        \textbf{E} Relationship across networks and stimuli between $\Delta \T{B}{early}$ and $\Delta \R{B}{}$, and associated linear regression.
    }
    \label{fig:early-adaptation}
    % 
    \figsupp[The equivalent figure in the full model with both TA and STD.]{%
        The equivalent of \FIG{early-adaptation} in the full model with both TA and STD.
        \textbf{A} The sample network, colored by latency rank, i.e. the rank order of the time to first spike in response to stimulation at site B in any sequence. The response develops asymmetrically due to structural properties, not due to interactions with A. Note that non-responding neurons are ranked last and appear dark blue.
        \textbf{B} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{B}{i}$, and associated linear regression. Neurons are colored by latency rank as in panel A. Additionally, excitatory neurons among the earliest 100 responders are shown in the left plot, while later responders are shown in the right plot. Inset histograms show the distributions of the plotted values, with the confidence intervals of their medians, calculated by bootstrap with 10000 iterations, in darker color.
        \textbf{C} Median non-target response contrast $\Delta \R{N}{}$ among the first 100 neurons to respond to stimulation in B, labeled ``early'', and among the later neurons, labeled ``late'', across networks and stimuli.
        \textbf{D} Median contrast in $\T{B}{}$ among the early and late neurons, respectively, across networks and stimuli.
        \textbf{E} Relationship across networks and stimuli between $\Delta \T{B}{early}$ and $\Delta \R{B}{}$, and associated linear regression.
    }
    {\includegraphics[width=\linewidth]{fig-early-adaptation-withSTD}}\label{figsupp:withSTD}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{fig-synergy}
    \caption{%
        In the full model, STD reduces \dev{} non-target and increases \dev{} target responses relative to \msc{}.
        \textbf{A} Average depression $\D{}{i}$ in target (A) trials (top row) and B trials (bottom row) in the \dev{} (left) and \msc{} sequence (center), and their contrast (right) in the sample network. The color scale is shared across columns (and between \dev{} and \msc{}) for direct comparability.
        \textbf{B} Relationship across neurons of the sample network between $\Delta \D{A}{i}$ and $\Delta \R{A}{i}$, and associated linear regression.
        \textbf{C} $\Delta \D{A}{}$ across networks and stimuli, plotted against the corresponding Pearson coefficient between $\Delta \D{A}{i}$ and $\Delta \R{A}{i}$.
        \textbf{D} $\Delta \D{B}{}$ in the first 100 neurons to respond to B, and in the remaining, late portion of the network, across networks and stimuli.
        \textbf{E} Relationship across networks and stimuli between $\Delta \D{B}{early}$ and $\Delta \R{B}{}$, and associated linear regression.
    }
    \label{fig:synergy}
    % 
    \figsupp[The equivalent figure in the ablated model with only STD.]{%
        The equivalent of \FIG{synergy} in the ablated model with only STD.
        \textbf{A} Average depression $\D{}{i}$ in target (A) trials (top row) and B trials (bottom row) in the \dev{} (left) and \msc{} sequence (center), and their contrast (right) in the sample network. The color scale is shared across columns (and between \dev{} and \msc{}) for direct comparability.
        \textbf{B} Relationship across neurons of the sample network between $\Delta \D{A}{i}$ and $\Delta \R{A}{i}$, and associated linear regression.
        \textbf{C} $\Delta \D{A}{}$ across networks and stimuli, plotted against the corresponding Pearson coefficient between $\Delta \D{A}{i}$ and $\Delta \R{A}{i}$.
        \textbf{D} $\Delta \D{B}{}$ in the first 100 neurons to respond to B, and in the remaining, late portion of the network, across networks and stimuli.
        \textbf{E} Relationship across networks and stimuli between $\Delta \D{B}{early}$ and $\Delta \R{B}{}$, and associated linear regression.
    }
    {\includegraphics[width=\linewidth]{fig-synergy-noTA}}\label{figsupp:noTA}
    %
    \figsupp[The STD variable $x_i$ only correlates with the activity of the immediately preceding trial.]{%
        The STD variable $x_i$ only correlates with the activity of the immediately preceding trial.
        Each marker represents the coefficient $\rho$ of the Pearson correlation between the spike counts in a target trial and the corresponding values $x_i$ at the beginning of the immediately following trial ($\Delta$ trials = 1) or after one intervening trial ($\Delta$ trials = 2). Recall that $x$ is reduced upon spiking, which implies that an effective correlation is negative. Color represents the p-value of the alternative hypothesis that $\rho < 0$. Horizontal lines indicate the median $\rho$ and p. Horizontal jitter is introduced to aid visual separability of the markers.
    }
    {\includegraphics[width=8cm]{sfig-STD-correlation-decay}}\label{figsupp:correlation-decay}
\end{figure}

@@

To show that $\D{}{}$ captures not only changes in $V_m$, by definition, but also in the resulting response, we correlated $\Delta \D{A}{i}$ with $\Delta \R{A}{i}$ across neurons of the sample network. Notice that this mirrors the correlation shown in \FIG{threshold-initial} between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$. As shown in \FIG{synergy}B, the relationship was clearly negative (Pearson's $\rho$ = -0.541, p = 2.28e-77; Wald test, n = 1000), indicating that higher $\D{}{}$ causes reduced responses as expected. We confirmed this relationship across networks and stimuli, see \FIG{synergy}C, and found that greater depression in \dev{} predicted a lower response in most networks (T = 724, p = 1.37e-14; one-sided Wilcoxon, n = 120). We note that calculating this same correlation in the model with STD only yields no significantly negative dependence (T = 3.99e+03, p = 0.828; see \FIGSUPP[synergy]{noTA}).

>>>>

    {\includegraphics[width=\linewidth]{fig-threshold-initial-withSTD}}\label{figsupp:threshold-initial-withSTD}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{fig-thresholds-follow-std}
    \caption{%
        Lower thresholds in \dev{} are a consequence of lower \dev{} non-target response magnitude.
        \textbf{A} $\T{A}{i}$ in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages in the \dev{} sequence ($\T[dev]{A}{i}$) and \msc{} sequence ($\T[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \T{A}{i}$). A and B stimulus locations are highlighted in the left plot; see \FIG{setup} for the locations of the remaining stimuli used in the \msc{} sequence.
        \textbf{B} Average response (in spikes per trial) to non-target trials ($\R{N}{i}$, with $N$ meaning B in \dev{}, and B, C, D, and E in \msc{}). Columns are arranged as in panel A.
        \textbf{C} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$, and associated linear regression.
        \textbf{D} $\Delta \R{N}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \R{N}{i}$ vs $\Delta \T{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
    }
    \label{fig:thresholds-follow-std}
    % 
    \figsupp[The equivalent figure in the full model with both TA and STD.]{%
        The equivalent of \FIG{thresholds-follow-std} in the full model with both TA and STD.
        \textbf{A} $\T{A}{i}$ in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages in the \dev{} sequence ($\T[dev]{A}{i}$) and \msc{} sequence ($\T[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \T{A}{i}$). A and B stimulus locations are highlighted in the left plot; see \FIG{setup} for the locations of the remaining stimuli used in the \msc{} sequence.
        \textbf{B} Average response (in spikes per trial) to non-target trials ($\R{N}{i}$, with $N$ meaning B in \dev{}, and B, C, D, and E in \msc{}). Columns are arranged as in panel A.
        \textbf{C} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$, and associated linear regression.
        \textbf{D} $\Delta \R{N}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \R{N}{i}$ vs $\Delta \T{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
    }
    {\includegraphics[width=\linewidth]{fig-thresholds-follow-std-withSTD}}\label{figsupp:thresholds-follow-std-withSTD}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{fig-early-adaptation}
    \caption{%
        Dev non-target responses are reduced due to adaptation in early responders to B.
        \textbf{A} The sample network, colored by latency rank, i.e. the rank order of the time to first spike in response to stimulation at site B in any sequence. The response develops asymmetrically due to structural properties, not due to interactions with A. Note that non-responding neurons are ranked last and appear dark blue.
        \textbf{B} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{B}{i}$, and associated linear regression. Neurons are colored by latency rank as in panel A. Additionally, excitatory neurons among the earliest 100 responders are shown in the left plot, while later responders are shown in the right plot. Inset histograms show the distributions of the plotted values, with the confidence intervals of their medians, calculated by bootstrap with 10000 iterations, in darker color.
        \textbf{C} Median non-target response contrast $\Delta \R{N}{}$ among the first 100 neurons to respond to stimulation in B, labeled ``early'', and among the later neurons, labeled ``late'', across networks and stimuli.
        \textbf{D} Median contrast in $\T{B}{}$ among the early and late neurons, respectively, across networks and stimuli.
        \textbf{E} Relationship across networks and stimuli between $\Delta \T{B}{early}$ and $\Delta \R{B}{}$, and associated linear regression.
    }
    \label{fig:early-adaptation}
    % 
    \figsupp[The equivalent figure in the full model with both TA and STD.]{%
        The equivalent of \FIG{early-adaptation} in the full model with both TA and STD.
        \textbf{A} The sample network, colored by latency rank, i.e. the rank order of the time to first spike in response to stimulation at site B in any sequence. The response develops asymmetrically due to structural properties, not due to interactions with A. Note that non-responding neurons are ranked last and appear dark blue.
        \textbf{B} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{B}{i}$, and associated linear regression. Neurons are colored by latency rank as in panel A. Additionally, excitatory neurons among the earliest 100 responders are shown in the left plot, while later responders are shown in the right plot. Inset histograms show the distributions of the plotted values, with the confidence intervals of their medians, calculated by bootstrap with 10000 iterations, in darker color.
        \textbf{C} Median non-target response contrast $\Delta \R{N}{}$ among the first 100 neurons to respond to stimulation in B, labeled ``early'', and among the later neurons, labeled ``late'', across networks and stimuli.
        \textbf{D} Median contrast in $\T{B}{}$ among the early and late neurons, respectively, across networks and stimuli.
        \textbf{E} Relationship across networks and stimuli between $\Delta \T{B}{early}$ and $\Delta \R{B}{}$, and associated linear regression.
    }
    {\includegraphics[width=\linewidth]{fig-early-adaptation-withSTD}}\label{figsupp:adaptation-withSTD}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=\linewidth]{fig-synergy}
    \caption{%
        In the full model, STD reduces \dev{} non-target and increases \dev{} target responses relative to \msc{}.
        \textbf{A} Average depression $\D{}{i}$ in target (A) trials (top row) and B trials (bottom row) in the \dev{} (left) and \msc{} sequence (center), and their contrast (right) in the sample network. The color scale is shared across columns (and between \dev{} and \msc{}) for direct comparability.
        \textbf{B} Relationship across neurons of the sample network between $\Delta \D{A}{i}$ and $\Delta \R{A}{i}$, and associated linear regression.
        \textbf{C} $\Delta \D{A}{}$ across networks and stimuli, plotted against the corresponding Pearson coefficient between $\Delta \D{A}{i}$ and $\Delta \R{A}{i}$.
        \textbf{D} $\Delta \D{B}{}$ in the first 100 neurons to respond to B, and in the remaining, late portion of the network, across networks and stimuli.
        \textbf{E} Relationship across networks and stimuli between $\Delta \D{B}{early}$ and $\Delta \R{B}{}$, and associated linear regression.
    }
    \label{fig:synergy}
    % 
    \figsupp[The equivalent figure in the ablated model with only STD.]{%
        The equivalent of \FIG{synergy} in the ablated model with only STD.
        \textbf{A} Average depression $\D{}{i}$ in target (A) trials (top row) and B trials (bottom row) in the \dev{} (left) and \msc{} sequence (center), and their contrast (right) in the sample network. The color scale is shared across columns (and between \dev{} and \msc{}) for direct comparability.
        \textbf{B} Relationship across neurons of the sample network between $\Delta \D{A}{i}$ and $\Delta \R{A}{i}$, and associated linear regression.
        \textbf{C} $\Delta \D{A}{}$ across networks and stimuli, plotted against the corresponding Pearson coefficient between $\Delta \D{A}{i}$ and $\Delta \R{A}{i}$.
        \textbf{D} $\Delta \D{B}{}$ in the first 100 neurons to respond to B, and in the remaining, late portion of the network, across networks and stimuli.
        \textbf{E} Relationship across networks and stimuli between $\Delta \D{B}{early}$ and $\Delta \R{B}{}$, and associated linear regression.
    }
    {\includegraphics[width=\linewidth]{fig-synergy-noTA}}\label{figsupp:synergy-noTA}
    %
    \figsupp[The STD variable $x_i$ only correlates with the activity of the immediately preceding trial.]{%
        The STD variable $x_i$ only correlates with the activity of the immediately preceding trial.
        Each marker represents the coefficient $\rho$ of the Pearson correlation between the spike counts in a target trial and the corresponding values $x_i$ at the beginning of the immediately following trial ($\Delta$ trials = 1) or after one intervening trial ($\Delta$ trials = 2). Recall that $x$ is reduced upon spiking, which implies that an effective correlation is negative. Color represents the p-value of the alternative hypothesis that $\rho < 0$. Horizontal lines indicate the median $\rho$ and p. Horizontal jitter is introduced to aid visual separability of the markers.
    }
    {\includegraphics[width=8cm]{sfig-STD-correlation-decay}}\label{figsupp:correlation-decay}
\end{figure}

@@

To show that $\D{}{}$ captures not only changes in $V_m$, by definition, but also in the resulting response, we correlated $\Delta \D{A}{i}$ with $\Delta \R{A}{i}$ across neurons of the sample network. Notice that this mirrors the correlation shown in \FIG{threshold-initial} between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$. As shown in \FIG{synergy}B, the relationship was clearly negative (Pearson's $\rho$ = -0.541, p = 2.28e-77; Wald test, n = 1000), indicating that higher $\D{}{}$ causes reduced responses as expected. We confirmed this relationship across networks and stimuli, see \FIG{synergy}C, and found that greater depression in \dev{} predicted a lower response in most networks (T = 724, p = 1.37e-14; one-sided Wilcoxon, n = 120). We note that calculating this same correlation in the model with STD only yields no significantly negative dependence (T = 3.99e+03, p = 0.828; see \FIGSUPP[synergy]{synergy-noTA}).
